---
title: "211019_2018_abundanc_demo"
author: "Lynette Lee"
date: "10/19/2021"
output: github_document
---

# Goal

This document shows how **individual bottle** bacterial abundance data from 2018 remineralization experiments were processed. Qc'd and analyzed. It also provides an intro to data processing and analysis with Rstudio and R Markdown. 

#Intro to R Markdown

Create a new code chunk: 
  Macs - Cmd + option + i
  Pc - Ctrl + alt + i

Load packages that we'll need to analyze our data. 

Run whole code chunk: 
  Macs - Cmd + shift + return 
  
```{r Load Packages, warning=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

We can toggle on/off warnings in chunks if we don't want them in our final markdown file. 

# Import Data

Macs: 
  <-: Option + -
  #: use for commenting, tells Rstudio not to run code
  
Direct file path
  tab completion: Tab + Command or Option
  
Go into Dataframe (from spreadsheet): use $ 

Join 2 dataframes together: 
  
```{r}
excel_sheets("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx")

#Mac: to make an arrow, use Option + - 

metadata <- read_excel("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Metadata")
glimpse(metadata)
unique(metadata$Bottle)
unique(metadata$Treatment)

data <- read_excel("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Data")
glimpse(data)

joined <- left_join(metadata, data) 
#attach data to metadata
#joins right dataset to the left on by using variables that aqre the same across the two dataframes
glimpse(joined)
#when joining data, must include the columns that are the same between the two dataframes

```

# Prepare Data 

We will convert the Date and Time column values from characters to dates, add columns with time elapsed for each treatment, and convert to cells/L because it will help us match up with the TOC data later. 

We will then subset the data for variables of interest and drop NA values.

To do this, we are going to use **piping**. Piping is an operation that allows us to write moer efficient code. The way that we'll use it here is to manipulate our data sequentially. 

Pipe operator "%>%", which basically says like "first do one thing to the data. THEN, don this other thing." (with the %>% operator taking the place of the word "THEN" in this scenario). 

Every call that we invoke with an additional pipe is going to take on the variable (dataframe) that we specify at the beginning.

```{r}
cells <- joined %>%
  mutate(Datetime = ymd_hm(Datetime), #splits apart Datetime as specified 
  cells_L = Cells_ml * 1000) %>%
  group_by(Treatment, Bottle) %>%
#group our dataset so that we can calculate the time elasped properly.
#calculate within the treatment, and within bottle, separately.
  mutate(interv = interval(first(Datetime), Datetime),
#calculate the interval Tn - T0
         s = as.numeric(interv),
         hours = s/3600,
         days = hours/24) %>%
  ungroup() %>%
  select(Experiment:DNA_Sample, cells_L, hours, days) %>%
#pipe knows that we are in "cell" and knows to select those column names.
#Experiment: DNA_Sample means everything in between Experiment to DNA_Sample column. 
  drop_na(cells_L)
glimpse(cells)
#drop the N/A values in the specific column named


```
