---
title: "211019_2018_abundanc_demo"
author: "Lynette Lee"
date: "10/19/2021"
output: github_document
---

# Goal

This document shows how **individual bottle** bacterial abundance data from 2018 remineralization experiments were processed. Qc'd and analyzed. It also provides an intro to data processing and analysis with Rstudio and R Markdown. 

#Intro to R Markdown

Create a new code chunk: 
  Macs - Cmd + option + i
  Pc - Ctrl + alt + i

Load packages that we'll need to analyze our data. 

Run whole code chunk: 
  Macs - Cmd + shift + return 
  
```{r Load Packages, warning=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

We can toggle on/off warnings in chunks if we don't want them in our final markdown file. 

# Import Data

Macs: 
  <-: Option + -
  #: use for commenting, tells Rstudio not to run code
  
Direct file path
  tab completion: Tab + Command or Option
  
Go into Dataframe (from spreadsheet): use $ 

Join 2 dataframes together: 
  
```{r}
excel_sheets("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx")

#Mac: to make an arrow, use Option + - 

metadata <- read_excel("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Metadata")
glimpse(metadata)
unique(metadata$Bottle)
unique(metadata$Treatment)

data <- read_excel("~/Desktop/GitHub/EEMB 144L/144l_student_2021/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Data")
glimpse(data)

joined <- left_join(metadata, data) 
#attach data to metadata
#joins right dataset to the left on by using variables that aqre the same across the two dataframes
glimpse(joined)
#when joining data, must include the columns that are the same between the two dataframes

```

# Prepare Data 

We will convert the Date and Time column values from characters to dates, add columns with time elapsed for each treatment, and convert to cells/L because it will help us match up with the TOC data later. 

We will then subset the data for variables of interest and drop NA values.

To do this, we are going to use **piping**. Piping is an operation that allows us to write moer efficient code. The way that we'll use it here is to manipulate our data sequentially. 

Pipe operator "%>%", which basically says like "first do one thing to the data. THEN, don this other thing." (with the %>% operator taking the place of the word "THEN" in this scenario). 

Every call that we invoke with an additional pipe is going to take on the variable (dataframe) that we specify at the beginning.

```{r}
cells <- joined %>%
  mutate(Datetime = ymd_hm(Datetime), #splits apart Datetime as specified 
  cells_L = Cells_ml * 1000) %>%
  group_by(Treatment, Bottle) %>%
#group our dataset so that we can calculate the time elasped properly.
#calculate within the treatment, and within bottle, separately.
  mutate(interv = interval(first(Datetime), Datetime),
#calculate the interval Tn - T0
         s = as.numeric(interv),
         hours = s/3600,
         days = hours/24) %>%
  ungroup() %>%
  select(Experiment:DNA_Sample, cells_L, hours, days) %>%
#pipe knows that we are in "cell" and knows to select those column names.
#Experiment: DNA_Sample means everything in between Experiment to DNA_Sample column. 
  drop_na(cells_L)
glimpse(cells)
#drop the N/A values in the specific column named


```

# Plot Growth Curves

We sill plot growth curve for each bottle. Need cell abundance and days data. 

First, let's set some aesthetics for our plot. 

```{r}
#assign hex colors to our different treatments
custom.colors <- c("Control" = "#377EB8", "Ash Leachate" = "#4DAF4A", "Mud Leachate" = "#E41A1C", "Glucose_Nitrate_Phosphate" = "#FF7F00")

#assign levels to control what order things appear in the legend
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

#now, let's use a handy package, ggplot to visualize our data. 

cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
#create a new column
#if DNA_Sample column is true, then new dna column is *
#if DNA_Sample column is not true, then new dna column is NA
  ggplot(aes(x=days, y=cells_L, group = interaction(Treatment, Bottle))) + 
#assign levels to control what order things appear in the legend
#lines connected by their respective Treatment and Bottles: group = interaction(Treatment, Bottle) 
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") +
  labs(x = "Days", y = expression(paste("Cells, L"^-1)), fill = " ") +
  guides(color = "none") +
#using custom colors assigned earlier
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  #facet_grid(rows = "Treatment")
#scale = "free": scale is not the same across the different graphs 
  theme_bw()

```
Increased growth in Ash Leachate and Glucose_Nitrate_Phosphate treatment.
Control had similar growth rate as the Mud Leachate treatments.

# Next Steps

We can caluclate: 

- total change in cells from initial conditions to the end of the experiment. 
- specific growth rate as the slope of ln(abundance) v time during exponential growth phase
- doubling time as ln(2) divide by the specific growth rate
- mean of each of these parameters across each treatment

1st, we need to determine **where** exponential growth in occurring in each of our bottles, if it actually does. To do this, we'll plot ln(abundance) vs time. 

# Identify exponential phase of growth of our remineralization experiment

**NOTE about logarithms in R**

log(x) gives the natural log of x, not log base 10
log10(x) gives the log base 10
log2(x) gives log base 2

```{r}
ln_cells <- cells %>%
#Ln_cells: dataframe
  group_by(Treatment, Bottle) %>%
  mutate(ln_cells = log(cells_L), 
#column name ln_cells
         diff_ln_cells = ln_cells - lag(ln_cells, default = first(ln_cells)))
#subtract ln_cell value from previous ln_cells value
#default = first(ln_cells) give zero value rather than NA

```

Now, let's plot our newly calcualted data!

```{r fig.height=6, fig.width=8}
#adjust dimensions of the graph
ln_cells %>%
  mutate(dna = ifelse (DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y= diff_ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1c") +
  labs(x = "Days", y = expression(paste("∆ln cells, L"^-1)), fill = " ") +
#Macs: delta is Option + J = ∆
  guides(color = "none") + 
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol = 2) +
#facet_wrap better than facet_grid when there is only one variable to subset by
#treatment is already specified by the colors
  theme_bw()
```
Exponential growth seems to be occurring right at the beginning of the experiment between 0-1 days for the most of the bottles. 

Let's try plotting ln_cells to see fi that can help us identify exponential growth in the control a little better. 

```{r fig.height=6, fig.width=8}
ln_cells %>%
  mutate(dna = ifelse (DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1c") +
  labs(x = "Days", y = expression(paste("ln cells, L"^-1)), fill = " ") +
  guides(color = "none") + 
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol = 2) +
  theme_bw()
```


