---
title: "211022_2021_Abundance_Homework"
author: "Lynette Lee"
date: "10/20/2021"
output: github_document
---
# Goal

Use the data collected of the inoculum from the flow cytometry to determine the growth curves of the bacterial communities over different treatments. The treatments include the control, which are filtered sea water; kelp exudate, which is the dissolved organic carbon (DOC) released from the macroalgae; kelp exudate supplemented with addition nutrients, which are nitrogen and phosphorous; and nutrient enriched sea water, where glucose, nitrogen, and phosphorous was added to the media. 


```{r warning=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

# Import Data

```{r}
excel_sheets("~/Desktop/GitHub/EEMB 144L/211019_2021_Abundance/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx")

metadata <- read_excel("~/Desktop/GitHub/EEMB 144L/211019_2021_Abundance/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Metadata")
#glimpse(metadata)

unique(metadata$Bottle)
unique(metadata$Treatment)

data <- read_excel("~/Desktop/GitHub/EEMB 144L/211019_2021_Abundance/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Data")
#glimpse(data)

joined <- left_join(metadata, data)
glimpse(joined)
```

# Prepare Data

The date and time data were separated.

The cell concentrations units were converted from total cells per microliters (uL) into total cells per liters (L).

```{r warning=FALSE}
cells <- joined %>%
  mutate(Datetime = ymd_hm(Datetime),
         all_cells_L = as.numeric(all_cells_uL) * 1000 * 1000,
         LNA_cells_L = as.numeric(LNA_cells_uL) * 1000 * 1000,
         HNA_cells_L = as.numeric(HNA_cells_uL) * 1000 * 1000) %>%
  group_by(Treatment, Bottle) %>%
  mutate(interv = interval(first(Datetime), Datetime),
         s = as.numeric(interv),
         hours = s/3600,
         days = hours/24) %>%
  ungroup() %>%
  select(Experiment:DNA_Sample, all_cells_L, hours, days) %>%
  drop_na(all_cells_L)
glimpse(cells)
```

# Plot Growth Curve

The growth curves of the microbial communities under different treatments over the span of 5 days were graphed together. The microbial growth peaked before the end of the 1st day for the control and kelp exudate treatments, while microbial growth peaked during the 2nd day for the supplemented kelp exudate and glucose-nitrate-phosphate treatments. The control also had similar growth curve to the kelp exudate treatmehnt. 

There were some missing time points since some of the labels on the FCM tubes fell off when storing them in liquid nitrogen. As a result, the data contained some NA values which were hidden earlier. 

```{r fig.height=6, fig.width=8}
custom.colors <- c("Control" = "#FAD7A0", "Kelp Exudate" = "#DAF7A6", "Kelp Exudate_Nitrate_Phosphate" = "#A3E4D7", "Glucose_Nitrate_Phosphate" = "#D7BDE2")

levels <- c("Control", "Kelp Exudate", "Kelp Exudate_Nitrate_Phosphate", "Glucose_Nitrate_Phosphate")

cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=all_cells_L, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) + 
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 10, color = "#F5B7B1") + 
  labs(x = "Days", y = expression(paste("Cells, L"^-1)), fill = " ") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  theme_bw()
```

# Calculations

To graph the natural log of the growth curve, the data was converted into the natural log values of the microbial cell concentrations. The incremental increase in cell concentration was found by subtracting the cellular concentration of Tn by the cellular concentration of Tn-1, which is the earlier time point.

```{r warning=FALSE}
ln_cells <- cells %>%
  group_by(Treatment, Bottle) %>%
  mutate(ln_cells = log(all_cells_L),
         diff_ln_cells = ln_cells - lag(all_cells_L, default = first(ln_cells)))
```

# Plotting Logarithmic Data

The logarithmic data was graphed and the data was separated by the different treatments and the bottles. The data shows the change in cellular concentration ovaer time. The exponential growth of the marine microbial communities appear to peak between the first treatment and day 1 for most treatment bottles. However, the kelp exudate with nitrate and phosphate treatment appears to have later exponential growth in between day 1 and 2. 

```{r fig.height=6, fig.width=8}
ln_cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=diff_ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 10, color = "#F5B7B1") +
  labs(x = "Days", y = expression(paste("∆ln cells, L"^-1)), fill = " ") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol = 2) +
  theme_bw()
```

# Identifying Exponential Growth

To graph the logarithmic graph of the cellular concentrations over time, the natural log values of the cellular concentrations were plotted. This graph better shows when the exponential growth occurred. 

Based on this data, the control and the kelp exudate treatment experienced exponential growth between days 0-1. The kelp exudate supplemented with nutrient treatment had one of its bottle showing exponential growth in between days 0 and 1, however one of its bottle has inconclusive data due to the NA value at T0. In addition, the glucose-nitrate-phosphate treatment also appears to have one of its data missing a T0 value while the other bottle displays exponential growth between T0 and T1.

```{r fig.height=6, fig.width=8}
ln_cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y=ln_cells, group = interaction(Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 10, color = "#F5B7B1") + 
  labs(x = "Days", y = expression(paste("ln cells, L"^-1)), fill = " ") +
  guides(color = "none") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol = 2) +
  theme_bw()
  
```

# Conclusion

The RStudio program efficiently gathers data from multiple sources and combines them, which was observed when the "metadata" was joined to "data." Then the newly joined data could be rearranged so that it can be later used to calculate new values and plot graphs. For example, when preparing the data for calculations, the datetime data points were revised so that the actual dates became seaparate from the time values. Once this new data was collected and labeled as "cells," this new data set was used to calculate natural log values to create logarithmic graphs. These logarithmic graphs better display the exponential growth of the marine microbial communities than simply observing change in cellular concentration. 


